---
title: "02: QC and reports"
author: "Andrew Lindsay, Reka Toth"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{02: QC and reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, knitr-setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE,
                      results= "hold",
                      eval=FALSE,
                      comment = "#>")

```

# Initial QC

Once the experiment is generated, we can create an interactive html report containing basic summary statistics of the [scMethrix] object with `scMethrix_report()` function. 

The report can be accessed here: [Initial reports](https://compepigen.github.io/WGBS_best_practice/initial_methrix_reports.html)

```{r}
methrix::methrix_report(meth = meth, output_dir = getwd(), prefix = "initial")
```

In the report, we can check genome-wide and chromosome based statistics on coverage and methylation, in order to identify potential quality issues, for example:

* samples with too high or too low genome-wide coverage or on a specific chromosome. In this case, there is also a possibility, that the sample is affected by copy number alterations. Both methylation levels and coverage therefore shows higher variation among cancer samples. 

* samples with altered beta level distribution. 

* A "bump" on the beta level density plot at 0.5 indicates the presence of single nucleotide polymorphisms (SNPs) -> see later.

# Filtering

## Masking low/high coverage CpGs 
To ensure the high quality of our dataset, the sites with very low coverage (untrustworthy methylation level) or high average coverage (technical problem) should not be used. We can mask these CpG sites. Some differential methylation region (DMR) callers, like DSS, will internally take into account low coverage sites, so it may not be necessary to mask them.

```{r}
# Masking CpG sites with a coverage < 2 and an average coverage > 2
scm <- scMethrix::mask_by_coverage(scm, low_threshold = 2, avg_threshold = 4) 
```

## Masking low sample count
If the coverage matrix is absent, filtering can be done via sample count instead. Similar to `mask_by_coverage`, `low_threshold` masks be low sample count, whereas `prop_threshold` will mask sites that appear in less than a certain proportion of all input samples.

```{r}
# Masking CpG sites present in fewer than 2 samples
scm <- scMethrix::mask_by_sample(scm, low_threshold = 2) 
```

```{r}
# Masking CpG sites present in fewer than 5% of samples
scm <- scMethrix::mask_by_sample(scm, prop_threshold = 0.05) 
```

## Masking low variance CpGs 
CpG sites with a low inter-individual methylation variance can increase the risk of false discoveries and reduce statistical significance. We can mask the sites to prevent this, as well as decreasing computation time and data size. The function `mask_by_variance` will calculate the variance of each CpG site and mask it if below a specified threshold. 

```{r}
# Masking sites with a variance < 0.05
scm <- scMethrix::mask_by_variance(scm, low_threshold = 0.05)
```

## SNP filtering

With SNPs, the C > T mutations can disrupt methylation calling. Therefore, it is essential to remove CpG sites that overlap with common variants, if we have variation in our study population. For example working with human, unpaired data, e.g. treated vs. untreated groups. 

During filtering, we can select the population of interest and the minimum minor allele frequency (MAF) of the SNPs.

SNP filtering is currently implemented for hg19 and hg38 assemblies. For mm10 assemblies, if the same strain of mice is used, SNP filtering is not generally necessary due to high homozygosity between individuals. SNP data for mice can be downloaded from here: https://www.sanger.ac.uk/science/data/mouse-genomes-project. 

```{r message=FALSE, warning=FALSE}
if(!requireNamespace("GenomicScores")) {
  BiocManager::install("GenomicScores")

}
library(GenomicScores) 
```

```{r}
scm <- scMethrix::remove_snps(scm, keep = TRUE)
```

* There is a possibility to keep the sites possibly overlapping with SNPs. We can use this data for detecting possible sample mismatches. For this, we need to plot the most variable sites (probably the ones that are variable in our study population) and plot them on a heatmap. 

```{r}
if (!requireNamespace("pheatmap", quietly = TRUE))
    install.packages("pheatmap")
snps <- meth[[2]]
meth <- meth[[1]]
pheatmap::pheatmap(get_matrix(order_by_sd(snps)[1:min(5000, length(snps)),]))
```

## Filtering masked sites

After masking is completed, we can remove sites that are no longer/never covered by any of the samples. 

```{r}
scm <- scMethrix::remove_uncovered(scm) 
```

# Querying

Often, whole genome analysis using all samples is not necessary. This package includes multiple convenient options for querying - `regions`, `contigs`, or `samples` - and can processed as either inclusive or exclusive. Multiple query types can be given in each command. Subset operations in [scMethrix] make use of the fast binary search `data.table` that is several orders faster than bsseq or other similar packages. 

```{r}
# Subsetting to include only two samples

scm.subset <- scMethrix::subset_scMethrix(scm, samples = gsm_list[1:2], by="include")
colData(scm.subset)
```

```{r}
# Subsetting to exclude two samples
scm.subset <- scMethrix::subset_scMethrix(scm, samples = gsm_list[1:2], by="exclude")
colnames(scm.subset)
```

```{r}
# Subsetting to include only chr1 and chr2
scm.subset <- scMethrix::subset_scMethrix(scm, contigs = c("chr1", "chr2"))
range(rowRanges(scm.subset))
```

```{r subset_scMethrix}
# Subsetting to include only bases 1:10M in chr5
regions <- GenomicRanges::GRanges(seqnames = "chr5", ranges = IRanges(1,10000000)) 
scm.subset <- scMethrix::subset_scMethrix(scm, regions = regions)
range(rowRanges(scm.subset))
```

# Visualization and QC after filtering

## scMethrix report
After filtering, it worth running a report again, to see if any of the samples were so lowly covered that it warrants an action (e.g. removal of the sample)

**Important:** Don't forget to set the `recal_stats` to `TRUE`, since the object changed since reading in. The output directory has to be different from last time, to avoid using the intermediate files calculated during the last run. 

```{r filtered_report}
methrix::methrix_report(meth = meth, output_dir = getwd(), recal_stats = TRUE, prefix="processed")
```
The report can be found here: [processed_methrix_reports.html](https://compepigen.github.io/WGBS_best_practice/processed_methrix_reports.html)

There are additional possibilities to visualize the study. We can look at the density of coverage both sample- and group-wise: 
  
```{r fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
scMethrix::plot_coverage(scm, type = "dens")
```

```{r fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
scMethrix::plot_coverage(scm, type = "dens", pheno = "Cell", show_legend = FALSE)
```

We can visualize te beta value distribution as a violin plot or density plot. These plots (as well as the coverage plot) use the 25000 most variable CpG sites to ensure fast processing. `plot_density` and `plot_violin` has the option to use data only from a restricted region. 

```{r fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
scMethrix::plot_density(scm, show_legend = FALSE)
```

```{r fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
scMethrix::plot_violin(scm, show_legend = FALSE)
```

[scMethrix]: https://compepigen.github.io/scMethrix/reference/scMethrix-class.html "`scMethrix`"