---
title: "04: Clustering and metacells"
author: "Andrew Lindsay, Reka Toth"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: true
toc_depth: 3
toc_float: true
self_contained: yes
highlight: pygments
vignette: >
  %\VignetteIndexEntry{04: Clustering and metacells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Clustering data
Unlike bulk sequencing, where the measurements are averaged over many cells, single-cell sequencing allows the identification of cell type for individual cells. This is a common step in many single-cell expression workflows. Multiple algorithms and distance metrics are included in the package for maximum flexibility. 

## Distance metrics
Most algorithms operate via a distance metric, an all-by-all matrix where each sample is numerically compared to each other sample. Samples with few methylated CpG sites in common will have a large distance metric, where as two samples with identical methylation will have a distance metric of zero. 

Available distance metrics include `pearson`, `spearman`, and `tau` (all via `bioDist`); and `euclidean`, `maximum`, `manhattan`, `canberra`, `binary`, and `minkowski` (all via `stats::dist`). 

```{r, eval=F}
scMethrix::get_distance_matrix(scm, assay = "score" , type = "pearson")
```

Additional distance metrics can be used via an arbitrary function input. The input must the assay matrix and output must be an all-by-all matrix filled with distance values for each respective pair. It will be internally cast with `as.dist()` before computation.

```{r, eval=F}
fun <- function(mtx) dist(mtx, method = "euclidean")
scMethrix::get_distance_matrix(scm, assay = "score" , type = fun)
```

## Clustering
Clustering algorithms include `hierarchical` (via `stats::hclust`), `partition` (via `stats::kmeans`), and `model`-based (via `mclust::Mclust`). Distance metrics are not used for model-based clustering. The identified clusters will be stored in the `colData` slot of the output `scMethrix` object.

```{r, eval=F}
scm.cluster <- scMethrix::cluster_scMethrix(scm, assay = "score", dist = dist, n_clusters = 2, colname = "Heirarchical", type="hierarchical")
colData(scm.cluster)
```

Like the distance metric, an arbitrary clustering algorithm can be used. It must accept a `dist` object, and return a `data.frame` with two columns named "Sample" and "Cluster". The column "Cluster" will be renamed by the value in `colname` before returning, if given. 

```{r, eval=F}
fun <- function (dist) {
  fit <- stats::hclust(dist, method="ward.D")
  fit <- stats::cutree(fit, k=ncol(dist)/2)
  colData <- data.frame(Sample = names(fit), Cluster = fit)
  colData
}

scm.cluster <- scMethrix::cluster_scMethrix(scm, assay = "score", dist = dist, n_clusters = 2, colname = "Heir", type=fun)
colData(scm.cluster)
```

# Visualization

For visualizing the clustering, the generated colData can be used for annotation in dimensionality reduction plots.

```{r, eval=F}
scm <- scMethrix::dim_red_scMethrix(scm,type="PCA",top_var = 10000)
plot_dim_red(scm,type="PCA",shape_anno = "Cluster",color_anno="Cluster")
```